---
title: "Statin, Anticoagulant & PPI IAD for Practice 37"
format: html
editor: visual
---

# **Summary Analysis of the Inter-arrival Density (time between two consecutive prescriptions for a chosen drug to be issued) for Statin, Anticoagulant and Proton Pump Inhibitor (PPI's) prescriptions in GP practice number 37**

```{r, include=FALSE}
##---##Getting Started##-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
library(tidyverse)
library(dplyr)     
library(lubridate)  
library(ggplot2)

##---##Isolating Data---------------------------------------------------------------------------------------------------------------------------------------------------------------------------

statins <- read.csv("/Users/padraicdonoghue/Desktop/GitHub/DIAMOND/Prescription intervals/Code/merged_output.csv")

##---##Formatting Date and Adding Month/Year Covariates-------------------------------------------------------------------------------

statins <- statins %>% 
  filter(str_detect(atc_final, "C10AA")) %>% 
  mutate(
    script_date = as.Date(as.character(script_date), format = "%d%b%Y"),
    prescription_month = month(script_date, label = TRUE, abbr = FALSE),  # Full month name
    prescription_year = year(script_date))

statins <- statins %>%
  arrange(UniquePatientID, script_date) %>%  # Critical for correct ordering
  group_by(UniquePatientID) %>%             # Essential for patient-specific gaps
  mutate(
    gap_days = as.numeric(difftime(lead(script_date), script_date, units = "days")),
    gap_days = abs(gap_days)  # Ensures positive values
  ) %>%
  ungroup()

##---Group for number of issues = 1 (30 days)---------------------------------------------
onemonth <- statins %>% 
  filter(numberofissues==1)

graphone <- ggplot(onemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 30 Day Statin Prescriptions Over A 1 Year Period",
    x = "Days Since Last Statin Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphone)

##---Group for number of issues = 3 (90 days)---------------------------------------------

threemonth <- statins %>% 
  filter(numberofissues==3)

graphtwo<- ggplot(threemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 90 Day Statin Prescriptions Over A 1 Year Period",
    x = "Days Since Last Statin Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphtwo)

##---Plotting------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

graphthree <- ggplot(statins, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between Statin Prescriptions Over A 1 Year Period",
    x = "Days Since Last Statin Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphthree)

##---Generating Simple Stats on the Histogram----------------------------------------------------------------------------------------------------------------------------------------------------------

summary(statins)

##---Generating Individual statin histograms-----------------------------------------------------------------------------------------------------------------------------------------------------------

multigraph<- ggplot(statins, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black", alpha = 0.6) +
  labs(
    title = "Density of Gaps Between Statin Prescriptions (by ATC Code)",
    x = "Days Since Last Statin Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  # facet_wrap(~ genericname) +  # Facet by statin ATC code/generic name
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot(multigraph)

```

# Statins

## This analysis is categorized to analyse differences in prescribing patterns and intervals across various healthcare scheme entitlements, genders and age brackets.

```{r}
library(knitr)
##---Statin Summary By ATC-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------
statin_summary_by_atc <- statins %>%
  # group_by(atc_final, genericname) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE)
  )
#view(statin_summary_by_atc)

##---Working with merged Data--------------------------------------------------------------------------------------------------------------------------------------------------------------------------

#Create age brackets
statins <- statins %>%
  mutate(age_bracket = case_when(
    age >= 60 & age <= 64 ~ "60–64",
    age >= 65 & age <= 69 ~ "65–69",
    age >= 70 & age <= 74 ~ "70–74",
    age >= 75 & age <= 79 ~ "75–79",
    age >= 80 & age <= 84 ~ "80–84",
    age >= 85 & age <= 90 ~ "85–90",
    age > 90 ~ "90+",
    TRUE ~ NA_character_
  ))

#Calculate gap_days by patient
statins <- statins %>%
  arrange(UniquePatientID, script_date) %>%
  group_by(UniquePatientID) %>%
  mutate(gap_days = as.numeric(script_date - lag(script_date))) %>%
  ungroup()

# Step 6: Group by sex,  scheme, age bracket
statin_summary <- statins %>%
  group_by(scheme, sex, age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )
kable(statin_summary, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins")

# Step 6: Group by scheme
statin_summary_scheme <- statins %>%
  group_by(scheme) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )
kable(statin_summary_scheme, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins by scheme")

# Step 6: Group by sex
statin_summary_sex <- statins %>%
  group_by(sex) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )
kable(statin_summary_sex, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins by sex")

# Step 6: Group by age bracket 
statin_summary_age_bracket <- statins %>%
  group_by(age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )
kable(statin_summary_age_bracket, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins by age bracket")

#group by year
statin_year <- statins %>%
  group_by(prescription_year) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )
kable(statin_year, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins by Year")

#group by month
statin_month <- statins %>%
  group_by(prescription_month) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = ifelse(count == 0, NA, mean(gap_days, na.rm = TRUE)),
    median_IAD = ifelse(count == 0, NA, median(gap_days, na.rm = TRUE)),
    sd_IAD = ifelse(count < 2, NA, sd(gap_days, na.rm = TRUE)), # SD needs ≥2 obs
    iqr_IAD = ifelse(count == 0, NA, IQR(gap_days, na.rm = TRUE)),
    min_IAD = ifelse(count == 0, NA, min(gap_days, na.rm = TRUE)),
    max_IAD = ifelse(count == 0, NA, max(gap_days, na.rm = TRUE)),
    .groups = "drop"
  )

kable(statin_month, caption = "Summary of Stastical Analysis of Prescribing Patterns for Statins by month")
```

## Below is an boxplot displaying the variation in results grouped by healthcare scheme and its associated standard deviation.

```{r}
#| fig.width: 10
#| fig.height: 6
ggplot(statins %>% filter(!is.na(gap_days)), 
                  aes(x = age_bracket, y = gap_days, 
                      fill = factor(sex),    # Fill color by sex
                      color = factor(sex))) + # Border color by sex
  geom_boxplot(alpha = 0.7) +  # alpha controls transparency
  facet_wrap(~ scheme) +
  scale_y_continuous(breaks = seq(0, max(statins$gap_days, na.rm = TRUE), by = 60)) +
  labs(
    title = "Distribution of Gap Days Between Statin Prescriptions",
    x = "Age Bracket",
    y = "Days Between Prescriptions",
    fill = "Sex",  # Legend title for fill
    color = "Sex"  # Legend title for border color
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Anticoagulants

```{r, include=FALSE}
##---##Getting Started##-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
library(tidyverse)
library(dplyr)     
library(lubridate)  
library(ggplot2)

##---##Isolating anticoagulants Data---------------------------------------------------------------------------------------------------------------------------------------------------------------------------

anticoagulants <- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/prescriptions_subsample_170625.csv")

anticoagulants <- anticoagulants %>% 
  mutate(atc_final = case_when(
    atc_final == "B01AA03" ~ "B01AZ",
    atc_final == "B01AB04" ~ "B01AZ",
    atc_final == "B01AB05" ~ "B01AZ",
    atc_final == "B01AB10" ~ "B01AZ",
    atc_final == "B01AE07" ~ "B01AZ",
    atc_final == "B01AF01" ~ "B01AZ",
    atc_final == "B01AF02" ~ "B01AZ",
    atc_final == "B01AF03" ~ "B01AZ",
    TRUE ~ atc_final
  ))

anticoagulants <- anticoagulants %>%
  filter(str_detect(atc_final, "B01AZ"))

##---##Formatting Date---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

anticoagulants <- anticoagulants %>% 
  mutate(script_date = as.Date(as.character(script_date), 
                               format = "%d%b%Y"))  

##---##Isolating one practice--------------------------------------------------------------------------------------------------------------------------------------------------------------------------

anticoagulantsGP37all <- anticoagulants %>% 
  filter(UniquePracticeID == 37) %>% 
  arrange(script_date) %>%
  mutate(gap_days = as.numeric(difftime(script_date, lag(script_date), units = "days")))

##---##Generating a column that includes the gap between previous and current RX-----------------------------------------------------------------------------------------------------------------------

anticoagulantsGP37all <- anticoagulantsGP37all %>%
  arrange(UniquePatientID, script_date) %>%  
  group_by(UniquePatientID) %>%
  mutate(gap_days = as.numeric(script_date - lag(script_date))) %>%
  ungroup()


##---Group for number of issues = 1 (30 days)---------------------------------------------
onemonth <- anticoagulantsGP37all %>% 
  filter(numberofissues==1)

graphone <- ggplot(onemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 30 Day anticoagulants Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last anticoagulants Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphone)

##---Group for number of issues = 3 (90 days)---------------------------------------------

threemonth <- anticoagulantsGP37all %>% 
  filter(numberofissues==3)

graphtwo<- ggplot(threemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 90 Day anticoagulants Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last anticoagulants Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphtwo)

##---Plotting------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

graphthree <- ggplot(anticoagulantsGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between anticoagulants Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last anticoagulants Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphthree)

##---Generating Simple Stats on the Histogram----------------------------------------------------------------------------------------------------------------------------------------------------------

summary(anticoagulantsGP37all)

##---Nice code that generates a line of best fit that could be good for presentations------------------------------------------------------------------------------------------------------------------
graphfour <- ggplot(anticoagulantsGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black", alpha = 0.6) +
  geom_density(colour = "darkred", size = 1) +
  labs(
    title = "Density of Gaps Between anticoagulants Prescriptions (Practice 37)",
    x = "Days Since Last anticoagulants Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphfour)

##---Generating Individual anticoagulants histograms-----------------------------------------------------------------------------------------------------------------------------------------------------------

multigraph<- ggplot(anticoagulantsGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black", alpha = 0.6) +
  labs(
    title = "Density of Gaps Between anticoagulants Prescriptions (by ATC Code)",
    x = "Days Since Last anticoagulants Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  facet_wrap(~ genericname) +  # Facet by anticoagulants ATC code/generic name
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot(multigraph)

##---anticoagulants Summary By ATC-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------
anticoagulants_summary_by_atc <- anticoagulantsGP37all %>%
  # group_by(atc_final, genericname) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE)
  )
#view(anticoagulants_summary_by_atc)

##---Merging Datasets----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# Step 1: Read in the data
demographics <- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/demographics_subsample.csv")
prescriptions <- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/prescriptions_subsample_170625.csv")
#view(prescriptions)
#view(demographics)
# Step 2: Merge the data on patient_id and practice_id
merged_data <- left_join(demographics, prescriptions, by = c("UniquePatientID", "UniquePracticeID"))

# Step 3: Optional - Save the merged data to a new CSV
write.csv(merged_data, "merged_output.csv", row.names = FALSE)
#view(merged_data)
```

## This analysis is categorized to analyse differences in prescribing patterns and intervals across various healthcare scheme entitlements, genders and age brackets.

```{r}
library(knitr)
##---Working with merged Data--------------------------------------------------------------------------------------------------------------------------------------------------------------------------
# Step 2: Filter anticoagulants

merged_data <- merged_data %>% 
  mutate(atc_final = case_when(
    atc_final == "B01AA03" ~ "B01AZ",
    atc_final == "B01AB04" ~ "B01AZ",
    atc_final == "B01AB05" ~ "B01AZ",
    atc_final == "B01AB10" ~ "B01AZ",
    atc_final == "B01AE07" ~ "B01AZ",
    atc_final == "B01AF01" ~ "B01AZ",
    atc_final == "B01AF02" ~ "B01AZ",
    atc_final == "B01AF03" ~ "B01AZ",
    TRUE ~ atc_final
  ))

anticoagulants_all <- merged_data %>%
  filter(str_detect(atc_final, "B01AZ"))

# Step 3: Convert script_date to Date type
anticoagulants_all <- anticoagulants_all %>%
  mutate(script_date = as.Date(as.character(script_date), format = "%d%b%Y"))
#view (anticoagulants_all)

# Step 4: Create age brackets
anticoagulants_all <- anticoagulants_all %>%
  mutate(age_bracket = case_when(
    age >= 60 & age <= 64 ~ "60–64",
    age >= 65 & age <= 69 ~ "65–69",
    age >= 70 & age <= 74 ~ "70–74",
    age >= 75 & age <= 79 ~ "75–79",
    age >= 80 & age <= 84 ~ "80–84",
    age >= 85 & age <= 90 ~ "85–90",
    age > 90 ~ "90+",
    TRUE ~ NA_character_
  ))

# Step 5: Calculate gap_days by patient
anticoagulants_all <- anticoagulants_all %>%
  arrange(UniquePatientID, script_date) %>%
  group_by(UniquePatientID) %>%
  mutate(gap_days = as.numeric(script_date - lag(script_date))) %>%
  ungroup()

# Step 6: Group by scheme, sex, age bracket — and summarise
anticoagulants_summary <- anticoagulants_all %>%
  group_by(scheme, sex, age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )

# Step 6: Group by scheme
anticoagulants_summary_scheme <- anticoagulants_all %>%
  group_by(scheme) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(anticoagulants_summary_scheme, caption = "Summary of Stastical Analysis of Prescribing Patterns for Anticoagulants by scheme")

# Step 6: Group by sex
anticoagulants_summary_sex <- anticoagulants_all %>%
  group_by(sex) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(anticoagulants_summary_sex, caption = "Summary of Stastical Analysis of Prescribing Patterns for Anticoagulants by sex")

# Step 6: Group by age bracket 
anticoagulants_summary_age_bracket <- anticoagulants_all %>%
  group_by(age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(anticoagulants_summary_age_bracket, caption = "Summary of Stastical Analysis of Prescribing Patterns for Anticoagulants by age bracket")
```

## Below is an boxplot displaying the variation in results grouped by healthcare scheme and its associated standard deviation.

```{r}
#| fig.width: 10
#| fig.height: 6
ggplot(anticoagulants_all %>% filter(!is.na(gap_days)), 
       aes(x = age_bracket, y = gap_days, 
           fill = factor(sex),    # Fill color by sex
           color = factor(sex))) + # Border color by sex
  geom_boxplot(alpha = 0.7) +  # alpha controls transparency
  facet_wrap(~ scheme) +
  scale_y_continuous(breaks = seq(0, max(anticoagulants_all$gap_days, na.rm = TRUE), by = 60)) +
  labs(
    title = "Distribution of Gap Days Between Anticoagulant Prescriptions",
    x = "Age Bracket",
    y = "Days Between Prescriptions",
    fill = "Sex",  # Legend title for fill
    color = "Sex"  # Legend title for border color
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Proton Pump Inhibitors (PPI's)

```{r, include=FALSE}
##---##Getting Started##-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
library(tidyverse)
library(dplyr)     
library(lubridate)  
library(ggplot2)

##---##Isolating PPI Data---------------------------------------------------------------------------------------------------------------------------------------------------------------------------

pp<- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/prescriptions_subsample_170625.csv")
PPI <- pp %>% 
  filter(str_detect(atc_final, "A02BC"))

##---##Formatting Date---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

PPI <- PPI %>% 
  mutate(script_date = as.Date(as.character(script_date), 
                               format = "%d%b%Y"))  

##---##Isolating one practice--------------------------------------------------------------------------------------------------------------------------------------------------------------------------

PPIGP37all <- PPI %>% 
  filter(UniquePracticeID == 37) %>% 
  arrange(script_date) %>%
  mutate(gap_days = as.numeric(difftime(script_date, lag(script_date), units = "days")))

##---##Generating a column that includes the gap between previous and current RX-----------------------------------------------------------------------------------------------------------------------

PPIGP37all <- PPIGP37all %>%
  arrange(UniquePatientID, script_date) %>%  
  group_by(UniquePatientID) %>%
  mutate(gap_days = as.numeric(script_date - lag(script_date))) %>%
  ungroup()


##---Group for number of issues = 1 (30 days)---------------------------------------------
onemonth <- PPIGP37all %>% 
  filter(numberofissues==1)

graphone <- ggplot(onemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 30 Day PPI Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last PPI Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphone)

##---Group for number of issues = 3 (90 days)---------------------------------------------

threemonth <- PPIGP37all %>% 
  filter(numberofissues==3)

graphtwo<- ggplot(threemonth, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between 90 Day PPI Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last PPI Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphtwo)

##---Plotting------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

graphthree <- ggplot(PPIGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black") +
  labs(
    title = "Density of Gaps Between PPI Prescriptions (Practice 37) Over A 1 Year Period",
    x = "Days Since Last PPI Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphthree)

##---Generating Simple Stats on the Histogram----------------------------------------------------------------------------------------------------------------------------------------------------------

summary(PPIGP37all)

##---Nice code that generates a line of best fit that could be good for presentations------------------------------------------------------------------------------------------------------------------
graphfour <- ggplot(PPIGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black", alpha = 0.6) +
  geom_density(colour = "darkred", size = 1) +
  labs(
    title = "Density of Gaps Between PPI Prescriptions (Practice 37)",
    x = "Days Since Last PPI Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  theme_minimal()
plot(graphfour)

##---Generating Individual PPI histograms-----------------------------------------------------------------------------------------------------------------------------------------------------------

multigraph<- ggplot(PPIGP37all, aes(x = gap_days, y = ..density..)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "black", alpha = 0.6) +
  labs(
    title = "Density of Gaps Between PPI Prescriptions (by ATC Code)",
    x = "Days Since Last PPI Prescription",
    y = "Density"
  ) +
  scale_x_continuous(breaks = seq(0, 365, by = 30)) +
  coord_cartesian(xlim = c(0, 365)) +
  facet_wrap(~ genericname) +  # Facet by PPI ATC code/generic name
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot(multigraph)

##---PPI Summary By ATC-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------
PPI_summary_by_atc <- PPIGP37all %>%
  # group_by(atc_final, genericname) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE)
  )
#view(PPI_summary_by_atc)






##---Merging Datasets----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# Step 1: Read in the data
demographics <- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/demographics_subsample.csv")
prescriptions <- read.csv("/Users/padraicdonoghue/Library/CloudStorage/OneDrive-SharedLibraries-RoyalCollegeofSurgeonsinIreland/Frank Moriarty - RSS 2025/prescriptions_subsample_170625.csv")
#view(prescriptions)
#view(demographics)
# Step 2: Merge the data on patient_id and practice_id
merged_data <- left_join(demographics, prescriptions, by = c("UniquePatientID", "UniquePracticeID"))

# Step 3: Optional - Save the merged data to a new CSV
write.csv(merged_data, "merged_output.csv", row.names = FALSE)
#view(merged_data)
```

## This analysis is categorized to analyse differences in prescribing patterns and intervals across various healthcare scheme entitlements, genders and age brackets.

```{r}
library(knitr)
##---Working with merged Data--------------------------------------------------------------------------------------------------------------------------------------------------------------------------
# Step 2: Filter PPI
PPI_all <- merged_data %>%
  filter(str_detect(atc_final, "A02BC"))

# Step 3: Convert script_date to Date type
PPI_all <- PPI_all %>%
  mutate(script_date = as.Date(as.character(script_date), format = "%d%b%Y"))
#view (PPI_all)

# Step 4: Create age brackets
PPI_all <- PPI_all %>%
  mutate(age_bracket = case_when(
    age >= 60 & age <= 64 ~ "60–64",
    age >= 65 & age <= 69 ~ "65–69",
    age >= 70 & age <= 74 ~ "70–74",
    age >= 75 & age <= 79 ~ "75–79",
    age >= 80 & age <= 84 ~ "80–84",
    age >= 85 & age <= 90 ~ "85–90",
    age > 90 ~ "90+",
    TRUE ~ NA_character_
  ))

# Step 5: Calculate gap_days by patient
PPI_all <- PPI_all %>%
  arrange(UniquePatientID, script_date) %>%
  group_by(UniquePatientID) %>%
  mutate(gap_days = as.numeric(script_date - lag(script_date))) %>%
  ungroup()

# Step 6: Group by scheme, sex, age bracket — and summarise
PPI_summary <- PPI_all %>%
  group_by(scheme, sex, age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )

# Step 6: Group by scheme
PPI_summary_scheme <- PPI_all %>%
  group_by(scheme) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(PPI_summary_scheme, caption = "Summary of Stastical Analysis of Prescribing Patterns for Proton Pump Inhibotors (PPI's) by scheme")

# Step 6: Group by sex
PPI_summary_sex <- PPI_all %>%
  group_by(sex) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(PPI_summary_sex, caption = "Summary of Stastical Analysis of Prescribing Patterns for Proton Pump Inhibotors (PPI's) by sex")

# Step 6: Group by age bracket 
PPI_summary_age_bracket <- PPI_all %>%
  group_by(age_bracket) %>%
  summarise(
    count = sum(!is.na(gap_days)),
    mean_IAD = mean(gap_days, na.rm = TRUE),
    median_IAD = median(gap_days, na.rm = TRUE),
    sd_IAD = sd(gap_days, na.rm = TRUE),
    iqr_IAD = IQR(gap_days, na.rm = TRUE),
    min_IAD = min(gap_days, na.rm = TRUE),
    max_IAD = max(gap_days, na.rm = TRUE),
    .groups = "drop"
  )
kable(PPI_summary_age_bracket, caption = "Summary of Stastical Analysis of Prescribing Patterns for Proton Pump Inhibotors (PPI's) by age bracket")

```

## Below is an error bar plot displaying the variation in results grouped by healthcare scheme and its associated standard deviation.

```{r}
#| fig.width: 10
#| fig.height: 6
ggplot(PPI_all %>% filter(!is.na(gap_days)), 
                  aes(x = age_bracket, y = gap_days, 
                      fill = factor(sex),    # Fill color by sex
                      color = factor(sex))) + # Border color by sex
  geom_boxplot(alpha = 0.7) +  # alpha controls transparency
  facet_wrap(~ scheme) +
  scale_y_continuous(breaks = seq(0, max(PPI_all$gap_days, na.rm = TRUE), by = 60)) +
  labs(
    title = "Distribution of Gap Days Between PPI Prescriptions",
    x = "Age Bracket",
    y = "Days Between Prescriptions",
    fill = "Sex",  # Legend title for fill
    color = "Sex"  # Legend title for border color
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
